---
title: "DA5030 Naive Bayes"
author: "Larkin, Yvonne"
date: "Fall 2025, October 14, 2025"
output:
  html_document:
    df_print: paged
---

```{r load_packages, warning=FALSE, message=FALSE}
library(knitr)
library(caret) #confusion matrix
library(e1071) #naive bayes
```

## Q2. Load CSV file
```{r Q2_load_csv}
url <- "http://s3.us-east-2.amazonaws.com/artificium.us/datasets/HouseVotes-1984-ButOne.csv"
#all categorical variables: Class Republican/Democrat and 17 legislatures vote cast for Features y/n
df <- read.csv(url, 
               header = T, 
               stringsAsFactors = T) # factors for naive bayes
head(df, 5)
```
## Q3. Get 5 lists of index split values
Desired target is 350 cases for training set, due to rounding have the p = (349/434) in `caret` function `createDataPartition`. I did not use `createFolds` due to the even split nature of *folds* but I also did not use `trainControl()` with `train()` due to the instructions specifying use of `e1071` package and its function `naiveBayes( )` to train the models. So these 5 manually split data partitions are "random" not "cross-validated".
```{r Q3_five_index_lists}
# 5 random sets to implement for testing
set.seed(932)

# split data and converted to df from lists
# select 349 because of uneven division value for desired vector length 350
x.split_index <- as.data.frame(createDataPartition(df$Class, 
                                                   p = (349/434), 
                                                   list = TRUE, 
                                                   times = 5))
```

## Q3. Split Data into 5 Training and 5 Testing Sets
```{r Q3_five_random_samples}
# training sets
df2.tr <- models_train.list <- lapply(1:ncol(x.split_index),
                      function(x){
                        # index training set
                        data.train <- df[x.split_index[[x]], ]
                      })
# testing set
df2.test <- models_test.list <- lapply(1:ncol(x.split_index),
                      function(x){
                        # index testing set
                        data.test <- df[-x.split_index[[x]], ]
                      })
```

## Q4. List of 5 Models
Use `e1071::naiveBayes( )` to trail each model with corresponding sample list. 
```{r Q4_naive_bayes_models}
# all 5 Naive Bayes models ; frequency tables used to predict
nb_models <- lapply(1:length(models_train.list), function(x){
  # access list of sample data ; extracted at data.frame level
  data <- models_train.list[[x]]
  # build  naive bayes classifiers ; classified feature 'Class'
  naiveBayes(subset(data, select = -Class), 
             data$Class, 
             laplace = 1)
})
```

## Q5. Build Ensemble Function `predict_part(models, new_cases)`
models = list ; new_cases = data.frame
A list of models and a data frame of Cases (including the "Class" column being Predicted) ccan be plugged into the **Emsemble Function** `predict_party(models_list, new_cases_df)` to generate and return the "Class" prediction for each case as a single vector that is the majority vote for all models used to evaluate the new case.
```{r Q5_ensemble_function}
# models as lists
# new_cases as df row
predict_party <- function(models, df.new_cases){
  # loop models and 
  for(m in 1:length(models)){
    # evaluate each row individually for all 5 models#new_cases <- 
    # make predictions and assign as list
    df.internal <- as.data.frame(predict(models[[m]], 
                                subset(df.new_cases, select = -Class)))
    # initialize counts
    Dcount <- Rcount <- 0
    #initialize empty vector 
    final_predict <- character(length = nrow(df.internal))
    # evaluate each row
    for(r in 1:nrow(df.internal)){
      Rcount <- sum(df.internal[r,] == 'republican', na.rm = TRUE)
      Dcount <- sum(df.internal[r,] == 'democrat', na.rm = TRUE)
      #assign majority
      final_predict[r] <- ifelse(Rcount == Dcount, "tie", ifelse(Rcount > Dcount, 'republican', 'democrat'))
      }
    final_predict
  }
  # Have to make sure to return list all the way outside the loops
  final_predict
  } 
```

## Q6. Calculate Accuracy Using Validation set
Here I am testing to make sure that the test sets do not result in a "tie" which could occur due to the present of `NA`s in votes. If "tie" did occur then a random selection for *'republican'* or *'democrat'* could be used if this feature needed to be used to assess further down the model, or for use in a different model. This is not necessary at this time and therefore I will leave "tie" as is.
```{r Q6_ensemble_model_accuracy, warning=FALSE}
#initialize prediction list
test_prediction <- vector("list", length = length(models_test.list))
# validation test
for(m in 1:length(models_test.list)){
  test_prediction[m] <- predict(nb_models, subset(models_test.list[[m]], select = -Class))
  ifelse(any(test_prediction[m] == 'tie') == TRUE, print(paste("Tie Value in list ", m)),next)
}
```

Loop through all 5 test samples generated from Q3 and use all models from Q4 to plug into the **Emsemble Function** `predict_party(model_list, new_case_df)` <br>
Saved all `confusionMatrix(table())` results in vector for visual manual exploration, but appended all table values to vectors to make a data frame from. Included Kappa as it is an adjusted Accuracy and therefore important when assessing the Accuracy decimal percent along with True Positive and True Negatives.

```{r Q6_find_overall_model_accuracy}
# initialize empty vectors
Model.Accuracy <- numeric(length = length(models_test.list))
Model.Kappa <- numeric(length = length(models_test.list))
True.Positive <- numeric(length = length(models_test.list))
True.Negative <- numeric(length = length(models_test.list))
# accuracy of each model
display.table <- table_accuracy <- vector("list", length = length(models_test.list))

# loop for confusionMatrix of each ensemble prediction run, against original sample run and extract desired display table
for(i in 1:length(models_test.list)){
  #save entire Confusion Matrix for evaluation outside loop ; set poisitive to 'democrat' to make sure all samples evaluated the same way
  table_accuracy[[i]] <- cM <- confusionMatrix(table(test_prediction[[i]], models_test.list[[i]][['Class']]), positive = 'democrat')
  
  # extract from current iteration by index of confusion Matrix
  # use double bracket to extract value only ; doesn't ask for the 
  Model.Accuracy[i] <- cM$overall[['Accuracy']]
  Model.Kappa[i] <- cM$overall[['Kappa']]
  True.Positive[i] <- cM$byClass[['Sensitivity']]
  True.Negative[i] <- cM$byClass[['Specificity']]
}

```

Format information for markdown display. Set the positive factor to 'democrat' to assure no switch in accuracy output sample to sample depending on the starting factor. 
```{r Q6_accuracy_Table}
display.df <- data.frame(row.names = c("Sample_1", "Sample_2", "Sample_3", "Sample_4", "Sample_5"),
                         Accuracy = Model.Accuracy,
                         Kappa = Model.Kappa,
                         True.Positive = True.Positive,
                         True.Negative = True.Negative)
#add the average of all 5 runs
display.df[nrow(display.df) + 1, ] <- sapply(display.df, mean)
row.names(display.df)[6] <- "Overall_Average"
# display table for markdown
display.md <- kable(round(display.df, 2), caption = "Ensemble Model's Accuracy Comparison by Labeled Test Set Run with 'democrat' as Positive Factor", )
```
`r display.md`

## Q7. New Case Predication
Appended new case to existing data frame so that the corresponding labels are assigned to the factor used. Use the `tail( )` to display and for `predict_party( )` function.
```{r Q7_Make_guess_party}
# define new case as data.frame
new_case <- c(NA, 'n', 'y', 'n', 'n', 'y', 'y', 'y', 'n', 'n', 'y', 'n', 'y', 'y', 'y', NA, 'n')
df.new_case <- rbind(df, new_case)
tail(df.new_case, 3)
```

```{r Q7_prediction}
# assign new case to variable (chose 'k' at random)
k <- tail(df.new_case, 1)
new_case.predict <- predict_party(nb_models, k)
new_case.predict
```

### Based on the Ensemble Function Model, the new case is predicted to be: **`r new_case.predict`**