---
title: "DA5030 Descriptive Analytics"
author: "Larkin, Yvonne"
date: "Fall 2025"
output:
  html_document:
    df_print: paged
---

### Q2
```{r Q2_load_csv_to_df, echo = FALSE}
#load kable libraries for table
library(knitr)
library(kableExtra)

# load csv from URL to df.wines
url <- "http://s3.us-east-2.amazonaws.com/artificium.us/datasets/whitewines.csv"
df.wines <- read.csv(url,
                     header = T,
                     # not necessary because all values numeric but assignment said to do so
                     stringsAsFactors = F) 
```

### Q3 
## Analysis of the Structure of the Data
```{r Q3_precheck_for_missing values, echo = F, include = FALSE}
# manually/visually checked for any missing values, there were none so I did not use na.rm parameter in calculations
missing_sugar <- which(is.na(df.wines$residual.sugar))
missing_alcohol <- which(is.na(df.wines$alcohol))
missing_sulphates <- which(is.na(df.wines$sulphates))
missing_quality <- which(is.na(df.wines$quality))
```

```{r Q3_calculate_stats, echo=FALSE}
#define my desired data frame by row name
df.results <- data.frame(row.names = c('residual sugar', 'alcohol', 'sulphates','quality'))

# calculate by row order for each columns value = median, 10% trim mean, sd [normality test separate]
# did not do this but in future will assign target to another variable to reduce typing/copy and reduce error potential
df.results$median <- sapply(df.wines[,c('residual.sugar','alcohol', 'sulphates', 'quality')], median)
df.results$mean <- sapply(df.wines[,c('residual.sugar','alcohol', 'sulphates', 'quality')], mean, trim = 0.1)
df.results$sd <- sapply(df.wines[,c('residual.sugar','alcohol', 'sulphates', 'quality')], sd)

#Shapiro-Wilks Tests for Normality -> Yes/No desired values
# APA citations and (long) explanation in 'Q3_citations_to_reduce_code' code chunk

# store (same ordered) normality tests in a list 
shapiro_list <- lapply(df.wines[,c('residual.sugar','alcohol', 'sulphates', 'quality')], shapiro.test)

# index each element of shapiro_list and assign actual p-values (not sublist of p.values) to a named vector
pvalues_vector <- sapply(shapiro_list, function(x) x$p.value)

# evaluate based on standard p-value cutoff of 0.05; sample is also < 5000 (observations = nrow in df.wines which is 4898)
# if p-value < 0.05 reject null hypothesis of normal distribution; meaning 'Not Normally Distributed'
df.results$normality <- ifelse(pvalues_vector < 0.05, 'No', 'Yes')


#APA Copilot help with <br> to format: Copilot. (2025, September 17). "kable() wrap text or line break". M365 Copilot.https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZESmtOMlU1TmpNM0xUWXdPVE10TkRjeFpTMDROelUxTFRSaU16aGtZelJsWWpNNE1Yd3lNREkxTFRBNUxURTNWREl6T2pRNE9qUXlMakEyTnprek5UbGEiLCJzY2VuYXJpbyI6InNoYXJlTGlua1ZpYVBvcG92ZXIiLCJwcm9wZXJ0aWVzIjp7InByb21wdFNvdXJjZSI6InVzZXIiLCJjbGlja1RpbWVzdGFtcCI6IjIwMjUtMDktMTdUMjM6NDg6NDIuMTgyWiJ9LCJjaGF0VHlwZSI6IndlYiIsInZlcnNpb24iOjEuMX0
# didn't actually ask to print this but since it was made I still printed it
Q3results_format <- kable(df.results, format = "html", digits = 2, caption = c("Analysis of Wine Data Metrics <br> and if Normally Distributed"), escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

`r Q3results_format`

### Q4 
## Correlation Analysis
```{r Q4_SpearmanRho_correlation_matrix, echo=FALSE, warning=FALSE}
# use Spearman Rho coefficient of correlation (method = 'spearman') to show the correlation between each pair of the features 
# need a 4x4 matrix ; correlation to self = 1, otherwise value will display twice because (row_alcohol x col_sulphates) == (col_alcohol x row_sulphates)

# initialize empty matrix instead of data frame because of assignment wording
spearman_matrix <- matrix(data = NA, 
                             nrow = 4, 
                             ncol = 4, 
                             dimnames = list(c('Residual Sugar','Alcohol', 'Sulphates', 'Quality'),
                                             c('Residual Sugar','Alcohol', 'Sulphates', 'Quality')))

# manually assigned them because things got complicated real fast defining as stand alone function and I was spending way too much time on it
row_metric <- c('residual.sugar','alcohol', 'sulphates', 'quality')
for(i in 1:4){
  spearman_matrix[i, 'Residual Sugar'] <- cor(df.wines[,row_metric[i]], df.wines$residual.sugar, method = 'spearman')
  spearman_matrix[i, 'Alcohol'] <- cor(df.wines[,row_metric[i]], df.wines$alcohol, method = 'spearman')
  spearman_matrix[i, 'Sulphates'] <- cor(df.wines[,row_metric[i]], df.wines$sulphates, method = 'spearman')
  spearman_matrix[i, 'Quality'] <- cor(df.wines[,row_metric[i]], df.wines$quality, method = 'spearman')
}

# format correlation matrix values in a table to print
Scor_table <- kable(spearman_matrix, digits = 2, format = "html", caption = "Spearman Rho Correlation Matrix <br> for Wine Metrics listed", escape = F) %>%
  kable_styling(bootstrap_options = "striped", full_width = F)
```

`r Scor_table`

### Q5
```{r Q5likem-coefficient_function, echo=FALSE}
# define function for likem-coefficient instead of vector operation for clarity
likem_coefficient <- function(a, s, h) {
  # a = alcohol, s = sugar, h = sulphates
  cal_value <- a / ((s^2)*sqrt(h))
  #log() in R is base e so ln() not base 10, defining base as 10 because that is math standard
  likem_co <- log(cal_value, base = 10)
}

# calculate new column for df.wines
df.wines$likem.coefficient <- likem_coefficient(df.wines$alcohol, df.wines$residual.sugar, df.wines$sulphates)

# find Spearman Rho correlation coefficient (to be consistent since none specified) to quality
qual_likem_cor <- cor(df.wines$quality, df.wines$likem.coefficient, method = 'spearman')

# evaluation for dynamic markdown printing ; made a function
cor_eval <- function(x){
  pos_neg <- ifelse(x > 0.0 & x != 0, "positive","negative")
  x <- abs(x)
  # initially used ifelse(x,'value', break | next | return) but first 2 broke loop and last overwrote variable relation, so did series of if(){}else if(){}
  if(x >= 0.0 & x < 0.25){
    relation <- paste('no relationship but', pos_neg)
    }else if(x >= 0.25 & x < 0.5){
      relation <- paste('weak', pos_neg)
      }else if(x >= 0.5 & x < 0.75){
        relation <- paste('moderate', pos_neg)
        }else if(x >= 0.75 & x < 0.99){
          relation <- paste('strong', pos_neg)
          }else if(x == 1){
            relation <- paste('perfect', pos_neg)
            }
  return(relation)
}
# relationship assessment of cor value between quality and likem-coefficient
relation_qlm <- cor_eval(qual_likem_cor)
```

*We define the "Likem-coefficient" by the formula below and found that there is a `r relation_qlm` correlation between the coefficient and "quality" (Spearman Rho coefficient of `r round(qual_likem_cor, digits = 2)`)*

$$
\log_{} \left( \frac{a}{s^{2} \sqrt{h}} \right)
$$
where, <br> <br>
*s* = "sugar", <br>
*h* = "sulphates", <br>
*a* = "alcohol" <br>

### Q6
```{r Q6_scatter_plot, echo=FALSE}
alc_sug_cor <- cor(df.wines$alcohol, df.wines$residual.sugar, method = 'spearman')

# basic scatterplot
plot(df.wines$alcohol,
     df.wines$residual.sugar,
     xlab = "Alcohol",
     ylab = "Residual Sugar",
     main = c("Scatterplot Correlation : ", "\n", "Between Wine's Alcohol and Sugar Content"),
     type = 'p')

# overlay general trend line used abline and lowess from R Help instructions just used the lowess for smoothed line
lines(lowess(df.wines$alcohol, df.wines$residual.sugar), col = 'orange')
```

From the scatter plot, it* visually appears that there are incremental steps down* in Residual Sugar Content as the Wine's Alcohol Content increases. The smooth lowess fit line does not have a steep slope but you can see it is a negative slope. *The Spearman Rho Coefficient* is equal to **`r round(alc_sug_cor, 2)`**, which agrees that there is **negative** correlation but also it would be categorized as a **weak correlation** between *alcohol* and *sugar*.

### Q7 
## Analysis of Sugar Content
```{r Q7_trim10_means, echo=FALSE}
#index to make 2 groupings
index_above10 <- which(df.wines$alcohol >= 10.0)
# object assign less10 and more10
less10 <- df.wines[-index_above10,'residual.sugar']
more10 <- df.wines[index_above10,'residual.sugar']

#indexed values for mean trim 10%
sugar_above10 <- mean(more10, trim = 0.10)
sugar_below10 <- mean(less10, trim = 0.10)

# difference in means use abs()
diff_mean10 <- abs(sugar_above10) - abs(sugar_below10)
abs_diff_mean <- abs(diff_mean10)

# test normality of each group's distribution
less10_shapiro <- shapiro.test(less10)
more10_shapiro <- shapiro.test(less10)

#if one OR both are not normally distributed use wilcox, if both are normal use t-test
wilcox_eval <- if(less10_shapiro$p.value < 0.05 | more10_shapiro$p.value < 0.05){
  
  # determine if only one is not normal
  num_not_norm <- ifelse(less10_shapiro$p.value < 0.05 & more10_shapiro$p.value < 0.05, "Both samples are not normally distributed ", "Only one sample is not not normally distributed ")
  
  # Wilcoxon Rank-Sum Test for not normal(non-parametric)
  #industry also uses z-test but not directly supported in R so using Wilcoxon rank-sum
  
  # determine if paired (Wilcoxon Sign) or unpaired (Wilcoxon Rank-Sum)
  logic_object <- ifelse(length(less10) == length(more10), TRUE, FALSE)
  
  # assign wilcoxon rank-sum test summary to object to index
  wilcox_alc_sugar <- wilcox.test(less10, more10, paired = logic_object, alternative = 'two.sided')
  
  # index pvalue to evaluate
  wilcox_pvalue <- wilcox_alc_sugar$p.value
  
  # indicate which Wilcoxon Test used cuz same function
  logic_object <- ifelse(logic_object == TRUE, " Paired Wilcoxon Sign", " Unpaired Wilcoxon Rank-Sum")
  
  #define a print statement based on 0.05 threshold for p.value
  wilcox_eval <- ifelse(wilcox_pvalue < 0.05, ". Because the p.value is less than alpha = 0.05, the null hypothesis is rejected. Meaning there is a ", ". Because the p.value is greater than alpha = 0.05, it fails to reject the null hypothesis, meaning there is no ")
  
  markdown_statement <- paste0(num_not_norm, "and the", logic_object, " Test calculated a p-value rounded to the fourth decimal place as ", format(round(wilcox_pvalue, 4), nsmall = 4, scientific = F), wilcox_eval, "statistically significant difference between the two sample means.")
} else {
  #if both normal distributions proceed with regular 2-sample t-test
  ttest_alc_sugar <- t.test(less10, more10)
  ttest_pvalue  <- ttest_alc_sugar$p.value
  # print statement for markdown

  wilcox_eval <- ifelse(wilcox_pvalue < 0.05, "The p-value is < 0.05, because of this the null hypothesis is rejected, meaning there is a statistically significant difference between the sample means.", "The p-value is > 0.05, because of this it fails to reject the null hypothesis. This indicates there is no statistically significant difference between the 2 sample means.")
}
```

The **10% trimmed mean** of *residual sugar* for wines with an *alcohol content* at or **above 10** is **`r round(sugar_above10, 1)`**. While in comparison the 10% trimmed mean of residual sugar for wines with an alcohol content **below 10** is **`r round(sugar_below10, 1)`**. This is a mean difference of **`r round(abs_diff_mean, 1)`**. `r wilcox_eval`

### Q8 
## Identification of Outliers
```{r Q8_identify_outliers, echo=FALSE}
#quality chlorides and density outlier z-score > 2.0
# use normal distribution for 95% CI on 2 tailed test
prob_norm <- qnorm(0.975)

# columns of interest
calc_cat <- df.wines[,c('quality', 'chlorides', 'density')]

# calculate each z score = (x-mean)/sd
df.wines$z_quality <- (df.wines$quality - mean(df.wines$quality))/sd(df.wines$quality)
df.wines$z_chlorides <- (df.wines$chlorides - mean(df.wines$chlorides))/sd(df.wines$chlorides)
df.wines$z_density <- (df.wines$density - mean(df.wines$density))/sd(df.wines$density)

# sum how many are outside of z-score > +/- 2.0
# because which returns a vector find length for # of outliers
quality_outliers <- length(which(abs(df.wines$z_quality) > 2.0))
chlorides_outliers <- length(which(abs(df.wines$z_chlorides) > 2.0))
density_outliers <- length(which(abs(df.wines$z_density) > 2.0))

# table rows = category ; columns = mean, sd , #outliers sd, z-score # outliers, 
df.outliers <- data.frame('Avg.Metric' = sapply(calc_cat, function(x) round(mean(x), 2)),
                          'St.Dev.' = sapply(calc_cat, function(x) round(sd(x), 2)),
                          'Avg.Zscore' = sapply(df.wines[,c('z_quality', 'z_chlorides', 'z_density')], function(x) round(mean(abs(x)), 2)),
                          '# Outliers' = sapply(c(quality_outliers, chlorides_outliers, density_outliers), round, digit = 2))

#rename rows and columns
rownames(df.outliers) <- c("Quality", "Chlorides", "Density")
colnames(df.outliers) <- c("Average <br> Wine Metric", "Standard <br> Deviation", "Average Metric <br> Z-score", "# Outliers <br> Z-score > 2.0")
  
#format df.outliers to display in HTML (did that since we could choose one)
format_outliers <- kable(df.outliers,
                         format = "html",
                         align = 'c',
                         escape = F,
                         caption = "Number of Outliers for Wine Data by Metric") %>%
  kable_styling(full_width = F, 
                bootstrap_options = "striped")
```

`r format_outliers`

```{r Q3_citations_to_reduce_code, echo=FALSE, include=FALSE}
# initial exploration of storing shapiro.test to a list and how, initially used sapply but that stored all 4 components ('statistic', 'p.value', 'method', 'data.name') as separate lists and so I needed lapply; then the opposit happened when trying to index within the lists I had created

# Utilized a lot of exploration in Console, Environment, and Help windows within RStudio to read on functions lapply(), sapply(), ifelse(), and Error feedbacks - I figured out many ways to proceed but wanted to try and figure out best way and why. Eventually, even though I found ways to continue, I asked Copilot for help on 2 parts:

# 1. 
#Why was ifelse() not working like a vector operation when indexing in a list <- kinda answers based on that but still I was confused, it told me sticking with sapply or lapply was the way to go. So I needed a function which was a lot of trial on notations and taking the error feedbacks

# APA citation : Copilot. (2025, September 16). "Error in shapiro_list[[x]]". M365 Copilot.https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZHVXhaRGMyTnpSbUxUWTROVGd0TkRRME9TMWhZbVJoTFdZMFl6QXhPV0U0WVdObVlYd3lNREkxTFRBNUxURTJWREl3T2pNNE9qQTNMakU1TWpZNE1WbyUzZCIsInNjZW5hcmlvIjoic2hhcmVMaW5rVmlhUG9wb3ZlciIsInByb3BlcnRpZXMiOnsicHJvbXB0U291cmNlIjoidXNlciIsImNsaWNrVGltZXN0YW1wIjoiMjAyNS0wOS0xNlQyMDozODowNi41MDJaIn0sImNoYXRUeXBlIjoid2ViIiwidmVyc2lvbiI6MS4xfQ

# create function x['p.value'] in order to index from list using sapply 

# sapply and lapply both gave me a list object but I prefer format of sapply since it tries to provide a vector. I could further condense this using arguement 'simplify = T'. Tested in console and explored in environment tab. This was great and I liked that I could easily index from it (did multiple tests with sapply and lapply)

# initially function(x) x['p.value'] but while investigating mode()/R type I arrived at a 'list' or 'numeric' depending on multiple indexing, assignment, and function evaluations. I want 'numeric' so I can test p value threshold. I figured out which notations and indexing WERE different and which would give me 'numeric' but I didn't understand why so I asked Copilot why list[[x]]['p.value'] is not the same as list[[x]]$p.value even though "==" would indicate they are and says the first method would need to unlist() 

# APA citation : Copilot. (2025, September 16). "Index difference in pvalues_list[[1]] + ['p.value'] and $p.value". M365 Copilot. https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZHTmlNVEUxTW1SaExUSXlNRGd0TkdRMU1pMWlObU0yTFdKall6STRabUUyWVRsallYd3lNREkxTFRBNUxURTJWREl4T2pRd09qUXdMamc0TURrME56RmEiLCJzY2VuYXJpbyI6InNoYXJlTGlua1ZpYVBvcG92ZXIiLCJwcm9wZXJ0aWVzIjp7InByb21wdFNvdXJjZSI6InVzZXIiLCJjbGlja1RpbWVzdGFtcCI6IjIwMjUtMDktMTZUMjE6NDA6NDAuMjY2WiJ9LCJjaGF0VHlwZSI6IndlYiIsInZlcnNpb24iOjEuMX0

# so with all of this I reduced multiple lines and trials to sapply() using $p.value to make numeric vector
```

