---
title: "DA5030 Decision Trees"
author: "Larkin, Yvonne"
data: 'Fall 2025: October 21, 2025'
output:
  html_document:
    df_print: paged
---

## Q1 Set-up R Project
```{r load_package_dependencies, warning=FALSE, message=FALSE}
library(knitr) # kable
library(ggplot2) # Q3
library(caret) # Q4
library(rpart) # Q5
library(rpart.plot) # Q6
library(C50) # Q9
```

## Q2 Load CSV File

```{r Q2_load_csv_file}
# load local file by name
df <- read.csv2(file = "bank-term-deposit-marketing-full.csv",
                header = TRUE,
                stringsAsFactors = TRUE)
# print header to check it loaded
head(df, 5)
```

## Q3 Data Exploration
```{r Q3_data_exploration}
str(df)
summary(df)

# broad search for any missing values, none
missing_val <- ifelse(any(is.na(df)) == FALSE, "There are no Missing Values `NA` found in the `df` features.", "Missing Values found in `df`")
```
```{r Q3_pdays_exploration}
# exploration due to str and summary outputs
neg_pdays <- length(which(df$pdays == -1)) # 36954 (-1)
prop_neg_val_pdays <-  round(neg_pdays/nrow(df), 3) # proportion -1
pdays_and_y_no <- which(df$pdays == -1 & df$y == "no") # 33570 (no&-1)
diff_pdays_y <- length(neg_pdays) - length(pdays_and_y_no) # 3384 [-1 - (no&-1)]

pout_unk <- length(which(df$poutcome == "unknown")) # 36959 (unknown)
pday_pout_comp <- length(which(df$pdays == -1 & df$poutcome == "unknown")) # 36954 (unknown&-1)
diff_pout_pday <- pout_unk -neg_pdays # 5 [unknown - (-1)] 

pout_and_y <- length(which(df$y == "no" & df$poutcome == "unknown")) # 33573 (y&no)
diff_y_pout <- length(which(df$y == "no")) - pout_and_y # 6349 [no - (no&unknown)]
```

```{r Q3_markdown_variables}
#initialize count and 
count <- 0
x <- vector("numeric", length = ncol(df))
for(i in 1:ncol(df)){
  if(is.integer(df[,i])){
    x[i] <- length(unique(df[[i]]))
    count <- (count + 1)
  } else {
    x[i] <- NA
    count <- count
  }
  x
  count
}
# variables for dynamic markdown
int_feat <- which(!is.na(x))
int_feat_name <- colnames(df)[int_feat]
non_factor <- data.frame(Feature = int_feat_name,
                         Unique_Values = x[int_feat])
```
### General Impressions
Download file to project and load by file name, name unchanged from downloaded name suggestion. Data is a semi-colon (;) delimited csv file for `read.csv2( )` is the default implementation for `sep = ";"` with `r ncol(df)` columns and `r nrow(df)` row observations. There is a mixture of numeric and character features, loaded the data where `stringAsFactor = TRUE` since `factor` values are needed for model generation and Decision Tree Use.<br><br>

**Target variable is "y"** and indicates whether the customer subscribed to a term deposit product offered by the bank, levels are : **`r levels(df$y)`**. Of the `r count` `integer` features, `r length(which(x[int_feat] < 100))` features have less than 100 unique values, however most numeric values are expansive enough that *discretization*  doesn't seem beneficial; especially since Decision Trees can handle both types of features. <br><br>

Based on the `str( )` and `summary( )` outputs there does seem to be an imbalance in many of the features, where there is an apparent strong correlation between some parameters, like "poutcome" and "pdays" even when compared to the target "y" for "no" subscription. Graphing the data will give a better distribution overview. Regardless of the distribution though, Decision Trees can work with either p"parametric" or "non-parametric" data. 


### Q3 Missing Values
**`r missing_val`** <br><br>
Thankfully no missing values need to be handled in this dataset, I say this because there are `r count` numeric features and `r ncol(df) - count` factor features, so depending  on the column a missing value is it would likely need more advanced techniques (models) to impute an appropriate value for it. It is likely that modeling techniques would be sufficient to predict a value. Numeric values could use kNN or Linear Regression (again due to the apparent high correlation of features) and categorical features could use Naive Bayes. I don't think more advanced techniques like deep learning etc would be needed (especially since we haven't covered those in the course), but I do not believe a mean/median/mode approach would be sufficient. There is the potential for  pairwise deletion in calculation. But overall Decision Trees and C5.0 specifically are not highly sensitive to missing values so if it we only a few *Missing Completely at Random Values* (MCAR) then I think it  would be fine to not address them at all when proceeding with the Decision Tree Creation.

### Q3 Numeric Feature Histograms (Density/Proportion)
```{r Q3_numeric_feat_Visualization, warning=FALSE, message=FALSE}
# to compare data to a normal distribution need to use densities/probabilities not frequencies
for (i in int_feat) {
  # data column
  num_feat <- df[[i]]
  # numeric plotting with ggplot2 T.T
  np <- ggplot(data.frame(x = num_feat), 
              aes(x)) +
    # set bin width to half the sd value of the feature so it is easier to visualize empirical rule
    geom_histogram(aes(y = ..density..), 
                   binwidth = sd(num_feat)/2) +
    stat_function(fun = dnorm,
                  args = list(mean = mean(num_feat), sd = sd(num_feat)),
                  color = "red", 
                  size = 1) +
    ggtitle(colnames(df)[i]) +
    theme_minimal()
  
  print(np)
}
# citations at end of all graphs
```
Surprisingly, the histograms of numeric features generally appear to follow a normal distribution. There peaks tend to be much more extreme though, so instead of a the empirical rule of 68-95-99.7 [^1] :
| Range from Mean | Percentage of Data |
|-----------------|--------------------|
| ±1σ             | ~68%               |
| ±2σ             | ~95%               |
| ±3σ             | ~99.7%             |
The data actually appears to be more like 95-99.7-100 or even 99.7-100-100 just looking at the graphs without calculating the actual proportion of data for each standard deviation value 'bin'. But again Decision Trees do not make distribution assumptions, it is mainly to bring to mind a type of **imbalance in the numeric data**, where values outside of 1 standard deviation of the mean, sampled in the training data.

### Q3 Categorical Feature Barplots (Frequency)
```{r Q3_categorical_feat_Visualization}
# sub-dataframe of only categorical features
cat_feat <- df[,-int_feat]
for(i in 1:ncol(cat_feat)){
  # data column
  num_feat <- cat_feat[[i]]
  # categorical plotting with ggplot2 T.T
  cp <- ggplot(data.frame(x = num_feat), 
              aes(x)) +
    # now looking at frequencies for each category, for discrete values (bins) so bar graph
    geom_bar()+
    ggtitle(colnames(cat_feat)[i]) +
    theme_minimal()
  
  print(cp)
}  
# citations at end of all graphs
```

Looking at the proportions this way is way faster than all of the `which()` search counts and comparisons, granted the `which()` searches allow for `&` operand to link features versus just looking at majority categorization but this is initial exploration so even linking isn't concrete and a linear regression model would be more appropriate to evaluate and correlations, or at least `cor.test()`.<br><br>

But it does **display multiple imbalances** in the categorical distributions, especially for the target value "y", so this means splitting the data must take cross validation into account because while Decision Trees can deal with imbalances in data to some degree, if the training data does not have a feature criterion present then it would be used for decision node splits, feature split rank evaluation, and so could overlook an important sub-categorization split meaning it could be miss classified due to default handling of the new category (majority class, alternate splits, or treated as a missing value).[^2]


Citation conversation a separate html doc linked because Copilot doesn't share conversations like ChatGPT (apologies for not realizing sooner) :
[^1]: initial understanding for histogram density use and dnorm ChatGPT https://share.google/aimode/ryXilfGgO6Abu338Z dnorm
[^1]: ggplot2 cheatsheet, since I never remember how to use ggplot2 file:///E:/Computer%20remove%20before%20reset/Documents/Northeastern%20Documents/ggplot2_Rvisualization_cheatsheet.pdf
[^1]: ('aes() meaning/use') https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZHSTJPREU1Tnpsa0xXVmlZMlF0TkRnNE5pMDVNbVF4TFdZMU1XRTRPR1JpTVdabVlud3lNREkxTFRFd0xUSXpWREl3T2pNM09qTXhMak0yTWpneU1WbyUzZCIsInNjZW5hcmlvIjoic2hhcmVMaW5rVmlhUG9wb3ZlciIsInByb3BlcnRpZXMiOnsicHJvbXB0U291cmNlIjoidXNlciIsImNsaWNrVGltZXN0YW1wIjoiMjAyNS0xMC0yM1QyMDozNzozMS4wODVaIn0sImNoYXRUeXBlIjoid2ViIiwidmVyc2lvbiI6MS4xfQ
[^1]:  ..density.. for histogram density https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZERTFZekZpTjJKbUxUSXhPVGt0TkdOa015MDVaamczTFRGak1UWTRZems0WXpFME1Id3lNREkxTFRFd0xUSXpWREl3T2pNNE9qUTRMakF6TnpVME4xbyUzZCIsInNjZW5hcmlvIjoic2hhcmVMaW5rVmlhUG9wb3ZlciIsInByb3BlcnRpZXMiOnsicHJvbXB0U291cmNlIjoidXNlciIsImNsaWNrVGltZXN0YW1wIjoiMjAyNS0xMC0yM1QyMDozODo0Ny43NzVaIn0sImNoYXRUeXBlIjoid2ViIiwidmVyc2lvbiI6MS4xfQ

[^1]:("empirical rule")[https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZEVXdPREUyTnpOaUxXSmtPREl0TkRGa05pMDVPRGc1TFdObVptSTFaakUyWkdVNFpud3lNREkxTFRFd0xUSXpWREl4T2pBd09qTTFMamMyTlRZeU1qZGEiLCJzY2VuYXJpbyI6InNoYXJlTGlua1ZpYVBvcG92ZXIiLCJwcm9wZXJ0aWVzIjp7InByb21wdFNvdXJjZSI6InVzZXIiLCJjbGlja1RpbWVzdGFtcCI6IjIwMjUtMTAtMjNUMjE6MDA6MzUuNTQxWiJ9LCJjaGF0VHlwZSI6IndlYiIsInZlcnNpb24iOjEuMX0]
[^2]: ("category not present in decision tree training")[https://m365.cloud.microsoft/chat/entity1-d870f6cd-4aa5-4d42-9626-ab690c041429/eyJpZCI6IlZYTmxjbFl4ZkdoMGRIQnpPaTh2YzNWaWMzUnlZWFJsTFdsdWRDNXZabVpwWTJVdVkyOXRMM3hQU1VRNlpUQXhaV1pqT1dFdE9EWmxOQzAwWlRVMkxUZzFOMlF0WldaaU1qa3lNVEk0TTJSa2ZHVTJOalppTURrekxXTTJaRGt0TkRGaVppMWhPVEUyTFRZd056Z3dPR0prTURjNE1Id3lNREkxTFRFd0xUSXpWREl4T2pJek9qSTJMakkwTXpFM05WbyUzZCIsInNjZW5hcmlvIjoic2hhcmVMaW5rVmlhUG9wb3ZlciIsInByb3BlcnRpZXMiOnsicHJvbXB0U291cmNlIjoidXNlciIsImNsaWNrVGltZXN0YW1wIjoiMjAyNS0xMC0yM1QyMToyMzoyNS45NTdaIn0sImNoYXRUeXBlIjoid2ViIiwidmVyc2lvbiI6MS4xfQ]

## Q4 Splitting Data
```{r Q4_split_data}
# set seed for repeatability of results
set.seed(7)
# caret function for 80:20 split
training_index <- createDataPartition(df$y,
                                      p = 0.8,
                                      list = FALSE)
# split data
training_set <- df[training_index,]
testing_set <- df[-training_index,]
```

```{r Q4_eval_target_proportions}
# make tables to print in markdown
target.train.prop <- kable(round(prop.table(table(training_set$y)),3), caption = "Training Data Set's Proportion of Categorical Factors")
target.test.prop <- kable(round(prop.table(table(testing_set$y)),3), caption = "Testing Data Set's Proportion of Categorical Factors")
```

```{r Q4_eval_all_categorical_feat_prop}
# initialize list object
object <- vector("list", length = ncol(cat_feat))
# loop categorical features to eval proportions
for(x in 1:ncol(cat_feat)){
  name <- colnames(cat_feat)[x]
  train.prop <- prop.table(table(cat_feat[training_index, x]))
  test.prop <- prop.table(table(cat_feat[-training_index, x]))
  # nested named list
  object[[x]] <- I(list(cat = name, 
                        train = train.prop, 
                        test = test.prop))
}
# printed object to look at all categories of nested list; removed for knit
```

`caret::createDataPartition` is the specified way to sample the data using an 80:20 split for percent of data in training versus testing sets. The function `createDataPartition( )` does not have a cross validation method parameter for performing the split and the C5.0 function to create the Decision Tree model does not have a cross validation parameter either so the best we can do is probability table to test proportions of the target variable because ideally the proportions will be nearly equivalent or at least not significantly different.  I visually compared the proportion of all categorical features, and luckily all features are present in the training and testing data, and also have very similar proportion values. So training the Decision Tree shouldn't be an issue.

### Target Variable `df$y` Proportions
`r target.train.prop`
<br>
`r target.test.prop`

## Q5 `rpart` package Decision Tree Classifier and Gini Index
```{r Q5_rpart_classifier}
# target variable is already factor
# used RHelp ; especially example for the parms = ; and 8.3 Classification trees p 150 - 160 : https://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf
rpart_fit <- rpart(y ~.,
      data = df,
      subset = training_index, # p. 160
      parms = list(prior = c(0.8, 0.2), 
                   split = "gini"), 
      method = "class")
# only did base ask from Q5 really
rpart_fit
```
The `rpart` model object results in:


This section took a lot of trial and error to complete, including: referencing R Help window for "{rpart}" and "{rpart.plot}" definitions and examples, reading the CranR documentation on Classification Methods Chapter 8 [^4] , and "Plotting rpart trees with the rpart plot package" vignette [^3] linked in RHelp, the linked "rpart.object" page, the assignment Hints and explicit parameters given. Also lots of playing around with indexing the model to see what is actually returned.

[^4]: https://cran.r-project.org/doc/contrib/Krijnen-IntroBioInfStatistics.pdf
[^3]: file:///C:/Users/yvonn/AppData/Local/R/win-library/4.5/rpart.plot/doc/prp.pdf ; searched online to get https://cran.r-project.org/web/packages/rpart.plot/index.html

## Q6 Visualize `rpart` Decision Tree
```{r Q6_rpart_visual_DT}
# decision tree figure for rpart
test_predictions <- rpart.plot(rpart_fit, 
                               type = 4)
```
*This was a really cool feature to learn about for the `rpart` package*

```{r Q6_visualization_node_splits}
# display the split order and rank 
r_import_feat <- rpart_fit$variable.importance
```

The root node is the highest split/importance which is the feature `r names(r_import_feat[1])`
`r round(r_import_feat, 2)`

## Q7 Evaluate Validation data Set
```{r Q7_rpart_test_set}
# make prediction
test_predict <- rpart.predict(rpart_fit, 
              newdata = testing_set[,-which(names(testing_set) == "y")],
              type = "class")

# evaluate accuracy
r.AC <- confusionMatrix(test_predict, testing_set$y, positive = "yes")
r.AC
```
The overall `rpart` model accuracy can be seen above. <br><br> 
The "positive" class was set to `r r.AC$positive` because having customers "subscribe to the term deposit product offered by the bank" is the desired outcome from the bank's perspective. The Overall accuracy for correct subscription prediction of "`r r.AC$positive`" is rather low at `r round(r.AC$byClass['Sensitivity'],2)` and is considered about `r round(r.AC$overall['Kappa'], 3)*100`% better than guessing, as indicated by the "Kappa" value. <br><br>
However, the false positive (indicated by 1 - "Specificity") rate is approximately `r round((1-r.AC$byClass['Specificity']), 2)`, which is much lower than the false negative (indicated by 1 - "Sensitivity") rate of `r round(r.AC$byClass['Sensitivity'], 2)` and likely the more favorable bias. This is because then there would not be an over estimate the potential revenue the bank generates from people's "subscription" (I assume a fee is associated); which be extension gives more leeway in end of the year budget review if they end up making more than anticipated. It is actually a interesting bias, considering the very low proportion of "yes" data observations to train the model on. But having a more accurate model of the true revenue generated by subscriptions is still considered preferable in accounting and in case any marketing strategies are to be derived from the types of individuals who *do subscribe*. Even if the overall model accuracy rate does not increase (`r round(r.AC$overall['Accuracy'],3)`) and improved kappa value (`r round(r.AC$overall['Kappa'], 3)`) or sensitivity (true positive) value would be beneficial. <br>

## Q8 Hyperparameter "Complexity Parameter" (*cp*) and "maxdepth"
### pre-tuning "cp" "or maxDepth"
```{r Q8_hyper_cp}
# current cp values
pre.cp <- rpart_fit$cptable
plotcp(rpart_fit)
```
Based on the plot of the current `rpart` model I think raising the 'cp' value to "0.031" might be better or even the "0.019" where it begins to flatten out would be better. But I will check all plot values above the default `cp = 0.01` and not at decision tree size = 1 split could be interesting.

### Complexity Parameter(cp)
```{r Q8_cp_change}
# initialize data
cp_val <- c(0.1, 0.067,0.031, 0.019, 0.013)
cp_fit <- vector("list", length = length(cp_val))
name_cp <- vector("character", length = length(cp_val))

# loop models
for(i in 1: length(cp_val)){
  name_cp[i] <- paste("cp =", cp_val[i], collapse = " ")
  cp_fit[[i]] <- rpart(y ~.,
                     data = df,
                     subset = training_index,
                     parms = list(prior = c(0.8, 0.2),
                                  split = "gini"),
                     method = "class",
                     cp = cp_val[i])
}
```

### Complexity Parameter cross-validation error plots
```{r Q8_cp_plots, warning=FALSE, message=FALSE}
# plot all cp ; for loop to avoid "NULL" plot with lapply()
for(i in 1:length(cp_fit)){
  plotcp(cp_fit[[i]])
}
```

### Q8 Extract Accuracy Values for cp models
```{r Q8_cp.df_comparison}
# initialize list
cp_confM <- vector("list", length = length(cp_fit))
for(i in 1:length(cp_fit)){
  preds <- rpart.predict(cp_fit[[i]],
                           newdata = testing_set[,-which(names(testing_set) == "y")],
                           type = "class")
  cp_confM[[i]] <- confusionMatrix(preds, testing_set$y, positive = "yes")
}
# df to print
cp.AC <- lapply(1:length(cp_fit), function(x){
  y <- cp_confM[[x]]
  y$overall['Accuracy']
  })
cp.kappa <- lapply(1:length(cp_fit), function(x){
  y <- cp_confM[[x]]
  y$overall['Kappa']
  })
cp.sens <- lapply(1:length(cp_fit), function(x){
  y <- cp_confM[[x]]
  y$byClass['Sensitivity']
  })
cp.spes <- lapply(1:length(cp_fit), function(x){
  y <- cp_confM[[x]]
  y$byClass['Specificity']
  })
cp.FP <- lapply(1:length(cp_fit), function(x){
  abs(cp.spes[[x]] - 1)
})
cp.FN <- lapply(1:length(cp_fit), function(x){
  abs(cp.sens[[x]] - 1)
})

cp.df <- data.frame(Model = name_cp,
                    Overall.Acc = unlist(cp.AC),
                    Kappa = unlist(cp.kappa),
                    TP_rate = unlist(cp.sens),
                    TN_rate = unlist(cp.spes),
                    FP_rate = unlist(cp.FP),
                    FN_rate = unlist(cp.FN))
```

 
### "maxDepth" Parameter
```{r Q8_maxDepth, warning=FALSE, message=FALSE}
# initialize data
depth_val <- c(2:17)
depth_fit <- vector("list", length = length(cp_val))
name_depth <- vector("character", length = length(depth_val))

# loop models
for(i in 1:length(depth_val)){
  name_depth[i] <- paste("depth =", depth_val[i], collapse = " ")
  depth_fit[[i]] <- rpart(y ~.,
                     data = df,
                     subset = training_index,
                     parms = list(prior = c(0.8, 0.2),
                                  split = "gini"),
                     method = "class",
                     control = rpart.control(maxdepth = i))
}
```

### "maxDepth" Parameter cross-validation error plots
```{r Q8_depth_plots, warning=FALSE, message=FALSE}
# plot all depth models ; for loop to avoid "NULL" plot with lapply()
for(i in 1:length(depth_fit)){
  plotcp(depth_fit[[i]])
}
```

### Q8 Extract Accuracy Values for maxDepth models
```{r Q8_depth.df_accuracy}
# initialize list
depth_confM <- vector("list", length = length(cp_fit))
for(i in 1:length(depth_fit)){
  preds <- rpart.predict(depth_fit[[i]],
                           newdata = testing_set[,-which(names(testing_set) == "y")],
                           type = "class")
  depth_confM[[i]] <- confusionMatrix(preds, testing_set$y, positive = "yes")
}
# df to print
depth.AC <- lapply(1:length(depth_val), function(x){
  y <- depth_confM[[x]]
  y$overall['Accuracy']
  })
depth.kappa <- lapply(1:length(depth_val), function(x){
  y <- depth_confM[[x]]
  y$overall['Kappa']
  })
depth.sens <- lapply(1:length(depth_val), function(x){
  y <- depth_confM[[x]]
  y$byClass['Sensitivity']
  })
depth.spes <- lapply(1:length(depth_val), function(x){
  y <- depth_confM[[x]]
  y$byClass['Specificity']
  })
depth.FP <- lapply(1:length(depth_val), function(x){
  abs(depth.spes[[x]] - 1)
})
depth.FN <- lapply(1:length(depth_val), function(x){
  abs(depth.sens[[x]] - 1)
})

depth.df <- data.frame(Model = name_depth,
                    Overall.Acc = unlist(depth.AC),
                    Kappa = unlist(depth.kappa),
                    TP_rate = unlist(depth.sens),
                    TN_rate = unlist(depth.spes),
                    FP_rate = unlist(depth.FP),
                    FN_rate = unlist(depth.FN))
```

```{r Q8_combined_df_accuracies}
changes.df <- rbind(cp.df, depth.df)
kable(changes.df, caption = "`cp` and `maxDepth` Impact on Accuracy Measures")
```

Visually inspecting all of the 'cp' alterations (`r paste(cp_val, collapse = ", ")`) and 'maxDepth' adjustments (`r paste(depth_val, collapse = ", ")`) from the original model cp value = "0.01", there isn't really any improvement to any of the parameters for accuracy. We'll see if *maxDepth* improves anything.
 
## Q9 `C5.0` Decition Tree with same data
```{r Q9_c50_DT}
# Fit C5.0 model with boosting
c5_model <- C5.0(training_set[,-which(colnames(training_set) == "y")],
                 training_set$y,
                 trials = 20)  # Boosting with 20 trials
c5_model
```

## Q10 Evaluate Q9 C5.0 model `c5_model`
```{r Q10_eval_c50model}
# make prediction
c5test_predict <- predict(c5_model, 
                        newdata = testing_set[,-which(names(testing_set) == "y")])

# evaluate accuracy
c5.AC <- confusionMatrix(c5test_predict, testing_set$y, positive = "yes")
c5.AC
```
```{r rpart_v_c5_models}
# r.AC and c5.AC confusion Matrix variables
base_models.df <- data.frame(Model = c("rpart Original Model", "C5.0 Boosted Model"),
                    Overall.Acc = c(r.AC$overall['Accuracy'], c5.AC$overall['Accuracy']),
                    Kappa = c(r.AC$overall['Kappa'], c5.AC$overall['Kappa']),
                    TP_rate = c(r.AC$byClass['Sensitivity'], c5.AC$byClass['Sensitivity']),
                    TN_rate = c(r.AC$byClass['Specificity'], c5.AC$byClass['Specificity']),
                    FP_rate = (1 - c(r.AC$byClass['Specificity'], c5.AC$byClass['Specificity'])),
                    FN_rate = (1 - c(r.AC$byClass['Sensitivity'], c5.AC$byClass['Sensitivity'])))
# combined model
base_models.df
```

The overall `c5_model` is actually very similar to the `rpart_fit`model for accuracy, kappa, and biases.<br> 
The "positive" class was set to `r r.AC$positive` because having customers "subscribe to the term deposit product offered by the bank" is the desired outcome from the bank's perspective. The Overall accuracy for correct subscription prediction of "`r c5.AC$positive`" is rather low at `r round(c5.AC$byClass['Sensitivity'],2)` and is considered about `r round(c5.AC$overall['Kappa'], 3)*100`% better than guessing, as indicated by the "Kappa" value. <br>
However, the false positive (indicated by 1 - "Specificity") rate is approximately `r round((1-c5.AC$byClass['Specificity']), 2)`, which is much lower than the false negative (indicated by 1 - "Sensitivity") rate of `r round(c5.AC$byClass['Sensitivity'], 2)` and likely the more favorable bias. This is because then there would not be an over estimate the potential revenue the bank generates from people's "subscription" (I assume a fee is associated); which be extension gives more leeway in end of the year budget review if they end up making more than anticipated. It is actually a interesting bias, considering the very low proportion of "yes" data observations to train the model on. But having a more accurate model of the true revenue generated by subscriptions is still considered preferable in accounting and in case any marketing strategies are to be derived from the types of individuals who *do subscribe*. Even if the overall model accuracy rate does not increase (`r round(c5.AC$overall['Accuracy'],3)`) and improved kappa value (`r round(c5.AC$overall['Kappa'], 3)`) or sensitivity (true positive) value would be beneficial.

## Q11 Comments
As the "No Free Lunch" Theorem explains, there isn't necessarily a superior algorithm  that it's dependent on the data set. And in this case to algorithms that both create Decision Tree models are being utilized. Because of the it would be expected that they perform similarly , especially for the same data splits/ data set. Their logic is likely similar, and both seem fairly well adjusted as they are part of official RStudio (posit) packages with documentation in the "Help" tab and on Cran R documentation. You could argue the `rpart` model is better because the "False Negative" rate is lower, which again would make the bank underestimate the revenue generated instead of accounting for *move revenue generation* and then falling short of it. But if you look at the model comparison data frame, everything has a very small different, certainly not one that would seem significantly better or worse in either direction.